Manuscript

### 2025-03-28 Manuscript_MD_ScoreReLiability.qmd
update MS add appendix

To assess whether ES heterogeneity can be reduced by correcting for
score reliability, estimates of meta-analytic heterogeneity are compared
across two situations: first, heterogeneity of raw ES, as computed in
equation 1 is assessed. Secondly, ES are corrected for imperfect
reliability. Computing the heterogeneity of these corrected ES allows
for an assessment, whether the heterogeneity did indeed shrink compared
to the first situation, as predicted in W&D and H&S.

@@

### Results - Wich phenomena pass criterion 3)?

@@

In that case, some heterogeneity in $\sigma_T$ would remain, albeit
heterogeneity found in $\sigma_x$ was reduced. @eq-ratio_d0 essentially
demonstrates how mean and heterogeneity of MD, paired with mean and
heterogeneity of total score standard deviation $\sigma_x$ can be used
to estimate heterogeneity in raw ES $\delta_0$. Similarly, we can adjust
Equation @eq-ratio_d0 to describe how, instead of total score standard
deviation, mean and variance in true score standard deviation $\sigma_T$
affect heterogeneity in corrected ES $\delta$.

@@

Heterogeneity in ES even in direct replications, where experimental
factors are held as constant as possible, is already substantial
(Renkewitz, Fünderich, & Beinhauer, 2024). However, so far we have
demonstrated that in several scenarios, differences in score reliability
did not serve as appropriate moderators or explanations of ES
heterogeneity. Across most phenomena, we failed to reduce ES
heterogeneity. If anything, we even increased ES heterogeneity, albeit
to a very small degree. Generally, concerning the multi-site replication
projects or collaborative research projects, accounting for differences
in score reliability did not help us gain a better understanding of the
phenomena.

@@

The analytical arguments presented following Equation 17 rested on the
use of the delta-method, as Equation 17 contained parameters of the
score standard deviations’ distributions, while the $R_1$ and $R_2$
metrics were designed to discuss score variances. The delta-method rests
on a number of assumptions: (1) approximations derived from the
delta-method require for the transformed variable to follow a marginal
normal distribution (ref). Since the variable under discussion here is
the standard deviation, which is bounded to be larger than zero and has
a non-normal sampling distribution (ref), this assumption is most likely
violated; (2) the derivative of the transformation function needs to be
available. As the transformation function is essentially just the square
root, this assumption should hold up; (3) as the delta-method
approximates the parameters of the transformed variable using the
first-order Taylor-expansion, this implies that higher-order terms are
required to be negligible. In practice, this implies that the variance
of the untransformed variable needs to be small. However, as this entire
procedure discusses the influence of variance in standard deviations on
ES heterogeneity, introduced by differences in score reliability, this
assumption is probably violated as well. The violation of these
assumptions likely introduce a bias in estimates derived from these
approximations (ref). However, the inequality derived with use of the
delta-method is not used to actually derive any estimates, but to
discuss how an attenuation correction procedure can affect ES
heterogeneity. Therefore, we believe that the violation of these
assumptions is defensible in this case, as the analytical arguments
derived from the inequality should still hold up, nevertheless. \[add to
appendix!\] Additionally, we used the empirical examples available to
assess in how far estimates derived using the delta-method
approximations diverge from estimates derived without it. The
differences identified are rather miniscule at the third decimal,
reinforcing our belief that these assumption violations are defensible.

Similarly, all Equations following Equation 10 rest on the assumption
that mean difference ($MD$) and true or total score variance
($\sigma^2_T$ and $\sigma^2_X$) are independent. While this assumption
is in line with assumptions underlying CTT, it is by no means guaranteed
that it holds in these data-sets. If this assumption was violated,
whether the relationship was positive or negative, indicated by a
positive or negative covariance in Equation 10, would most likely have
an impact on how strongly ES heterogeneity is affected by the
attenuation correction procedure, possibly even the direction of how it
is affected. However, currently we have no reason to assume that there
is a systematic relationship between mean difference and standard
deviation. As this relationship would also affect the distribution of ES
like Cohen’s d itself, strong violations of this assumption would go far
beyond invalidating the claims we derived in Equation 16. Instead,
assumptions of the meta-analytic tests concerning the size ES and the
presence of ES heterogeneity would be violated, making any research on
change in ES heterogeneity due to attenuation correction obsolete.

@@

>>>>

To assess whether ES heterogeneity can be reduced by correcting for
score reliability, estimates of meta-analytic heterogeneity are compared
across two situations: first, heterogeneity of raw ES, as computed in
@eq-d is assessed. Secondly, ES are corrected for imperfect reliability.
Computing the heterogeneity of these corrected ES allows for an
assessment, whether the heterogeneity did indeed shrink compared to the
first situation, as predicted in W&D and H&S.

@@

### Results - Which phenomena pass criterion 3)?

@@

In that case, some heterogeneity in $\sigma_T$ would remain, albeit
heterogeneity found in $\sigma_x$ was reduced. @eq-ratio_d0 essentially
demonstrates how mean and heterogeneity of MD, paired with mean and
heterogeneity of total score standard deviation $\sigma_x$ can be used
to estimate heterogeneity in raw ES $\delta_0$. Similarly, we can adjust
@eq-ratio_d0 to describe how, instead of total score standard deviation,
mean and variance in true score standard deviation $\sigma_T$ affect
heterogeneity in corrected ES $\delta$.

@@

### Limitations & constraints on generalizability

@@

The analytical arguments leading to @eq-ineq_short rested on the use of
the delta-method, as @eq-ineq_init contained parameters of the score
standard deviations’ distributions, while the $R_1$ and $R_2$ metrics
were designed to discuss score variances. The delta-method rests on a
number of assumptions, at least one of which is likely violated in our
analysis: the variance of the variable to be transformed needs to be
small (ref.). As this entire procedure discusses the influence of
variance in standard deviations on ES heterogeneity, introduced by
differences in score reliability, assuming that this assumption holds
seems somewhat off. Therefore, we used the empirical examples available
to assess in how far estimates derived using the delta-method
approximations diverge from estimates derived without it. The
differences identified are rather miniscule for the majority of
phenomena, at the third decimal. Only for the PSACR002_behav_init
phenomenon we find larger differences - therefore, conclusions for this
phenomenon concerning the relation between $R_1$ and $R_2$ need to be
drawn with caution. Generally, as the majority of phenomena did not
display any bias in this assessment and since the delta-method is not
actually used to generate estimates, we believe its use to be defensible
nonetheless.

Similarly, all equations following @eq-XYZ_cov rest on the assumption
that mean difference ($MD$) and true or total score variance
($\sigma^2_T$ and $\sigma^2_X$) are independent. While this assumption
is in line with assumptions underlying CTT, it is by no means guaranteed
that it holds in these data-sets. If this assumption was violated,
whether the relationship was positive or negative, indicated by a
positive or negative covariance in @eq-XYZ_cov, would most likely have
an impact on how strongly ES heterogeneity is affected by the
attenuation correction procedure, possibly even the direction of how it
is affected. However, currently we have no reason to assume that there
is a systematic relationship between mean difference and standard
deviation. As this relationship would also affect the distribution of ES
like Cohen’s d itself, strong violations of this assumption would go far
beyond invalidating the claims we derived in @eq-ineq_short. Most
likely, assumptions of the meta-analytic tests concerning the
meta-analytic mean size ES and the presence of ES heterogeneity would be
violated, making any research on change in ES heterogeneity due to
attenuation correction obsolete.

### Conclusions

Heterogeneity in ES even in direct replications, where experimental
factors are held as constant as possible, is already substantial
(Renkewitz, Fünderich, & Beinhauer, 2024). However, so far we have
demonstrated that in several scenarios, differences in score reliability
did not serve as appropriate moderators or explanations of ES
heterogeneity. Across most phenomena, we failed to reduce ES
heterogeneity. If anything, we even increased ES heterogeneity, albeit
to a very small degree. Generally, concerning the multi-site replication
projects or collaborative research projects, accounting for differences
in score reliability did not help us gain a better understanding of the
phenomena. We want to reinforce that, in light of uncontrolled variation
in score variance due to the lack of appropriate sampling schemes,

@@

## Appendix

### A - Rearranging inequality

Both metrics $R_1$ and $R_2$ make use of parameters of the distribution
of true and total score variance, not standard deviations. This choice
was made to facilitate the role of score reliability in the subsequent
discussion, as score reliability itself is defined by variances, not
standard deviations (see @eq-rel).

Using the metrics $R_1$ and $R_2$, defined in [Equations @eq-R1] and
[-@eq-R2], we attempt to understand under which circumstances the
inequality defined in @eq-ineq_init holds true. However, while the
metrics make use of mean and heterogeneity of the score variances,
@eq-ineq_init makes use of mean and heterogeneity of the standard
deviations. In order to incorporate the metrics, it is necessary to
reparameterise the equation. Using the delta method to approximate mean
value and heterogeneity of variance from those of the standard
deviations, we know that (ref):
$$\mu^2_{\sigma_X} \approx \mu_{\sigma^2_X} \quad \quad  and \quad \quad 
\tau^2_{\sigma_X} \approx \frac{\tau^2_{\sigma_X^2}}{4\mu_{\sigma^2_X}}$$ {#eq-delta_method}

While @eq-delta_method only explicitly contains parameters for the
observed score variance/standard deviation, the same can be done using
the true score variance/standard deviation. Using the approximates
defined in @eq-delta_method for observed and true score variance, we
arrive at a new inequality in @eq-ineq_CV.

$$\frac{\tau_{MD}^2}{\mu_{\sigma^2_t}} + \frac{\mu_{MD}^2}{\mu^2_{\sigma_t^2}} \frac{\tau^2_{\sigma_t^2}}{4\mu_{\sigma^2_t}} > \frac{\tau_{MD}^2}{\mu_{\sigma^2_x}} + \frac{\mu_{MD}^2}{\mu^2_{\sigma_x^2}} \frac{\tau^2_{\sigma_x^2}}{4\mu_{\sigma^2_x}}$$ {#eq-ineq_CV}

@eq-ineq_CV describes that the heterogeneity in corrected ES is larger
than the heterogeneity in uncorrected ES, using parameters of the
distributions of mean difference MD, true score variance $\sigma^2_T$
and observed score variance $\sigma^2_X$. By introducing the metrics
$R_1$ and $R_2$ to this inequality, we can begin to disentangle which
conditions need to be fulfilled for the inequality to hold. Rearranging
the terms in [Equations @eq-R1] and [-@eq-R2], and subsequently entering
these terms into @eq-ineq_CV leads to the following
$$\frac{\tau_{MD}^2}{R_1 \mu_{\sigma^2_x}} + \frac{\mu_{MD}^2}{R_1^2 \mu^2_{\sigma^2_x}} \frac{R_2 \tau^2_{\sigma_x^2}}{4 R_1 \mu_{\sigma^2_x}} > \frac{\tau_{MD}^2}{\mu_{\sigma^2_x}} + \frac{\mu_{MD}^2}{\mu^2_{\sigma_x^2}} \frac{\tau^2_{\sigma_x^2}}{4\mu_{\sigma^2_x}}$$ {#eq-ineq_R}

@eq-ineq_R, again, describes the inequality that heterogeneity in
corrected ES is larger than the heterogeneity in uncorrected ES,
incorporating the metrics $R_1$ and $R_2$. This equation alone is not
sufficient to identify the relevant circumstances required for the
inequality to hold. However, rearranging the terms from @eq-ineq_R leads
to @eq-ineq_fin:

$$ \tau_{MD}^2\left(\frac{1}{R_1} - 1\right) + \mu_{MD}^2 \frac{1}{4} CV^2_{\sigma^2_X} \left(\frac{R_2}{R_1^3}  - 1\right) > 0$$ {#eq-ineq_fin}

The circumstances under which the inequality described in @eq-ineq_fin
holds, are circumstances where the claims made in W&D and H&S are
directly contradicted, as in those cases the ES heterogeneity is larger
after applying an attenuation correction procedure. Concerning the terms
in @eq-ineq_fin, we know that all terms left-hand of the inequality,
outside of the brackets ($\tau^2_{MD}$, $\mu^2_{MD}$ and
$CV^2_{\sigma^2_X}$), are bound to be positive. For the inequality to
hold, we need the left-hand side of the equation to remain positive,
larger than zero. We can distill two scenarios, under which this
equation should hold: (a) one of the terms inside the brackets
($\frac{1}{R_1} - 1$ and $\frac{R^2_2}{R_1}  - 1$) is positive and large
enough, so that the second term not containing that same bracket is
positively dominated by the first term; or (b) both terms inside the
brackets need to be positive.

Generally, we know that both $R_1$ and $R_2$ are bound to be positive,
as both contain different, strictly positive parameters of the
distributions of true and observed score variance. Additionally, we know
that $R_1$ is equivalent to the average score reliability, and therefore
bound between $0\le R_1\le1$. Therefore, the term inside the first
bracket is bound to be positive ($\frac{1}{R_1} – 1$). Similarly, $R_2$,
as the ratio of true and error score variance heterogeneity, is bound
between $0\le R_2\le1$. If $R_1^3$ is smaller than $R_2$, then the term
inside the second bracket turns negative. Generally, this means that the
inequality can only be violated if the reduction in relative
heterogeneity of score variance ($R_2$) is larger than the the average
score reliability to the power of 3. Restated in mathematical terms,
this leads us to the rule/condition described in @eq-ineq_short, in the
main text.

### B - Back-transformation log-transformed score variance

Using the methdology underlying the Boot-Err approach (Beinhauer et al.
ref), we generate estimates of meta-analytic mean and heterogeneity in
log-transformed total score variance $\sigma^2_X$. To facilitate
interpretation and allow for computation of the metrics $R_1$ and $R_2$,
we need to back-transform those estimates to the original scale. Using
@eq-mu_sigmaX we can generate an estimate of meta-analytic mean score
variance on the scale of $\sigma^2_X$ and using @eq-tau_sigmaX we get an
estimate of its heterogeneity.

$$\hat{\mu}_{\sigma^2_X} = exp\left(\hat{\mu}_{ln[\sigma^2_X]} + .5\  \hat{\tau}^2_{ln[\sigma^2_X]}\right)$$ {#eq-mu_sigmaX}

$$\hat{\tau^2}_{\sigma^2_X} = exp\left(\hat{\tau}^2_{ln[\sigma^2_X]} - 1\right) \cdot \exp\left(2\ \hat{\mu}_{ln[\sigma^2_X]} + \hat{\tau}^2_{ln[\sigma^2_X]}\right)$$ {#eq-tau_sigmaX}

In [equations @eq-mu_sigmaX] and [-@eq-tau_sigmaX],
$\hat{\mu}_{ln[\sigma^2_X]}$ refers to the mean estimate from the
meta-analysis on log-transformed score variance,
$\hat{\tau}^2_{ln[\sigma^2_X]}$ refers to the estimate of heterogeneity.

In [equations @eq-mu_sigmaE] and [-@eq-tau_sigmaE] we generate estimates
of meta-analytic mean and heterogeneity of error score variance
$\sigma^2_E$. Just like for the total score variance, we make use of the
estimates of mean and heterogeneity from the meta-analysis on
log-transformed error score variance.

$$\hat{\mu}_{\sigma^2_E} = exp\left(\hat{\mu}_{ln[\sigma^2_E]} + .5\  \hat{\tau}^2_{ln[\sigma^2_E]}\right)$$ {#eq-mu_sigmaE}

$$\hat{\tau^2}_{\sigma^2_E} = exp\left(\hat{\tau}^2_{ln[\sigma^2_E]} - 1\right) \cdot \exp\left(2\ \hat{\mu}_{ln[\sigma^2_E]} + \hat{\tau}^2_{ln[\sigma^2_E]}\right)$$ {#eq-tau_sigmaE}

### C - Examining bias due to delta-method

To examine whether applying the delta-method to derive the conditions
under which the inequality @eq-ineq_init holds, we compared estimates of
mean and heterogeneity of total score standard deviation$\sigma_X$. To
generate estimates that do not depend on use of the delta-method, we
used bootstrapping and a log-transformation, similar as detailed in
@sec-methods2 for the score variances $\sigma^2_X$. Using the natural
log, we transformed estimates of score standard deviation and using
bootstrapped samples we derived appropriate standard errors. From a
random-effects meta-analysis we generated estimates of mean and
heterogeneity and back-transformed those to the original scale of
$\sigma_X$. Additionally, we used the delta-method as outlined in
@eq-delta_method to estimate mean and heterogeneity in $\sigma_X$ from
estimates derived from meta-analyses on $\sigma^2_X$. The estimates can
be found in @tbl-delta.

```{r, echo = FALSE}
#| label: tbl-delta
#| tbl-cap: Empirical assessment of appropriateness of delta-method
knitr::kable(read.csv(here("Tables/delta_test.csv")),
             col.names = c("MASC", 
                           "$\\hat{\\mu}_{\\sigma_X}$", 
                           "$\\hat{\\tau}^2_{\\sigma_X}$", 
                           "$\\hat{\\mu}_{\\Delta \\sigma_X}$", 
                           "$\\hat{\\tau}^2_{\\Delta \\sigma_X}$"))
```

Generally, for 11 out of 12 phenomena, we find miniscule differences
between the two estimation strategies. For all but the
PSACR002_behav_init phenomenon, the differences occur in the 3rd
decimal, indicating that violations in the assumptions concerning the
delta-method are negligible here. Only for the PSACR002_behav_init
phenomenon we find larger differences. These are large enough to be
concerning and pertain largely the estimate of heterogeneity.
Accordingly, inferences regarding this phenomenon need to be interpreted
with caution. Generally, since for the majority of phenomena we find no
bias and since the delta-method is not used to generate estimates beyond
this particular section, we believe its use to be defensible.
